Racial Force Probe

### 2024-02-15 paper/paper.qmd
Data cleaning done and graphs completed

#### Workspace setup ####
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("here")
# install.packages("ggplot2")
# install.packages("janitor")
# install.packages("zoo")

library(ggplot2)
library(tidyverse)
library(knitr)
library(here)
library(janitor)
library(zoo)
```

@@

@@

```{r}
#| message: false
#| echo: false
#| label: fig-overall-death
#| fig-cap: Causes of deaths among the prisoners
```

@@

pop <- c(1217352, 412783, 328239523)  # population denominators

# Read CSV file with the correct date format
x <- read.csv("data/raw_data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
x$date <- as.Date(x$date, format = "%m/%d/%Y")  # Specify the correct date format

# Calculate rolling averages
pp <- 1e5 * zoo::rollapply((x$pp_cases[-1] - x$pp_cases[-132]), width = 7, FUN = mean) / pop[1]
sp <- 1e5 * zoo::rollapply((x$sp_cases[-1] - x$sp_cases[-132]), width = 7, FUN = mean) / pop[2]
gp <- 1e5 * zoo::rollapply((x$gp_cases[-1] - x$gp_cases[-132]), width = 7, FUN = mean) / pop[3]

# Combine data into a data frame
ts_data <- data.frame(
  date = rep(x$date[-(1:7)], 3),
  new_cases_per_100k = c(pp, sp, gp),
  group = rep(c("Incarcerated", "Prison Staff", "General Population"), each = length(pp))
)

@@

health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))

health_demographic_data_clean <-
  clean_names(health_data_count_by_demographics) |>
  select("x1", "cancer", "heart_disease", "liver_disease",
         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
         "drug_alcohol_intoxication", "accident", "homicide_d")

# Create a mapping for all causes
cause_mapping <- c(
  "cancer" = "Cancer",
  "heart_disease" = "Heart Disease",
  "liver_disease" = "Liver Disease",
  "respiratory_disease" = "Respiratory Disease",
  "aids_related" = "Aid related",
  "all_other_illnesses" = "All other illnesses",
  "suicide" = "Suicide",
  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
  "accident" = "Accident",
  "homicide_d" = "Homicide"
)

# Rename the causes in the dataset
health_demographic_data_clean <- health_demographic_data_clean %>%
  rename_with(~ cause_mapping[.], -x1)

gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))

gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")

# Calculate proportions within each category
gender_health_data_long <- gender_health_data_long %>%
  group_by(x1) %>%
  mutate(Proportion = Count / sum(Count))

# Plot the proportions graph using geom_bar with position = "fill"
ggplot(gender_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
  labs(title = "Proportions of Causes of Death by Category",
       x = "Gender",
       y = "Proportion") +
  scale_fill_discrete(labels = cause_mapping) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

# Read data
health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))

# Clean and organize data
health_demographic_data_clean <-
  clean_names(health_data_count_by_demographics) |>
  select("x1", "cancer", "heart_disease", "liver_disease",
         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
         "drug_alcohol_intoxication", "accident", "homicide_d")

# Pivot data to long format
health_demographic_data_long <- pivot_longer(health_demographic_data_clean, cols = -x1, names_to = "Cause", values_to = "Count")

# Filter data for specific demographics (e.g., "Male" and "Female")
filtered_df <- health_demographic_data_clean[health_demographic_data_clean$x1 %in% c("Male", "Female"),]

# Filter data for specific races (e.g., "White", "Black", "Hispanic", "Other")
race_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("White", "Black", "Hispanic", "Other"))

# Pivot data to long format for race
race_health_data_long <- pivot_longer(race_health_data, cols = -x1, names_to = "Cause", values_to = "Count")

# Calculate proportions instead of counts
race_health_data_long <- race_health_data_long %>%
  group_by(x1) %>%
  mutate(Proportion = Count / sum(Count))

# Create a mapping for all causes
cause_mapping <- c(
  "cancer" = "Cancer",
  "heart_disease" = "Heart Disease",
  "liver_disease" = "Liver Disease",
  "respiratory_disease" = "Respiratory Disease",
  "aids_related" = "Aid related",
  "all_other_illnesses" = "All other illnesses",
  "suicide" = "Suicide",
  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
  "accident" = "Accident",
  "homicide_d" = "Homicide"
)

# Rename the causes in the dataset
race_health_data_long <- race_health_data_long %>%
  mutate(Cause = cause_mapping[as.character(Cause)])

# Create a bar plot showing proportions with black borders
ggplot(race_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
  labs(title = "Proportions of Causes of Death by Category",
       x = "Category",
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

>>>>

#### Workspace setup ####
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("here")
# install.packages("ggplot2")
# install.packages("readxl")

library(ggplot2)
library(readxl)
library(tidyverse)
library(knitr)
library(here)
```

@@

## Source
For this analysis we have used multiple datasets. This paper uses the datasets available in the replication package of the original paper. The author of the original paper has collated a large range of files from multiple sources and we will be using some of the datasets from this collation.

## Methodology
For this paper, the datasets were cleaned and analysed using the statistical programming software R [@citeR] along with the help tidyverse [@citeTidyverse], knitr [@citeKnitr], ggplot2 [@citeGgplot], here [@citeHere], zoo [@citeZoo], readxl [@citeRead], writexl [@citeWrite] and janitor [@citeJanitor].

## Datasets
### mass-incarceration.xlsx

This dataset is a compilation of data collated by the original paper’s author from different sources:
- Sourcebook of Criminal Justice Statistics
- Bureau of Justice Statistics Report
- Jail Inmates in 2018
- U.S. Population FRED Data
- World Prison Population Data
- Western B. and Pettit B data

There are three sheets in this file, we will be using sheet 1 which contains time series data for incarceration from 1925-2019; each year is associated with the number of imprisoned individuals per 100k people, calculated from the total population and total number of inmates.
We will use this dataset to get the counts of prisons for different years. The Prison variable in the dataset is that variable.

### msfp0116stt02data.csv and msfp0116stt09data.csv
The original paper’s author got this data from
- the BJS publication by E. Ann Carson, Mary P. Cowhig (2020), Mortality in State And Federal Prisons
- 2001-2016 Statistical Tables (NCJ 251920)

msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016. These causes include blah blah blah. For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over

msfp0116stt09data.csv contains counts for the same conditions but segmented by demographic factors instead of time, such as gender, race, and age. For our analysis, we will be using the stats segmented by gender and race.

### covidts01.csv
The original paper’s author got this data from
- Covid Prison Project (https://covidprisonproject.com) by Lauren BrinkelyRubinstein (UNC Chapel Hill) and Kathryn Nowotny (University of Miami).

This dataset contains the collective number of new cases and case rates per 100,000 among the incarcerated, prison staff, and the general population. The numbers are recorded on a daily basis from April 22, 2020 to the end of August 2020.

We will now analyse these datasets to uncover trends and draw conclusions.

@@

```{r}
#| message: false
#| echo: false
#| label: fig-overall-death
#| fig-cap: Causes of deaths among the prisoners

# Load cleaned data from CSV
health_data_by_year_long_prop <- read.csv("data/analysis_data/cleaned_health_data_by_year.csv", header = TRUE, stringsAsFactors = FALSE)

# Create the plot
ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
  geom_line(size = 1.0) +
  labs(title = "Trends of Diseases by Year",
       x = "Year",
       y = "Proportion of cases",
       color = "Disease") +
  theme_minimal()

@@

# Load cleaned data
ts_data <- read.csv("data/analysis_data/cleaned_covid19_data.csv", header = TRUE, stringsAsFactors = FALSE)

@@

# Load cleaned data
health_demographic_data_clean <- read_csv("data/analysis_data/cleaned_health_data_gender.csv")

# Plot the proportions graph using geom_bar with position = "dodge"
ggplot(health_demographic_data_clean, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
  labs(title = "Proportions of Causes of Death by Category",
       x = "Gender",
       y = "Proportion") +
  scale_fill_discrete(labels = cause_mapping) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

# Load cleaned data from CSV
race_health_data_long <- read.csv("data/analysis_data/cleaned_health_data_race.csv", header = TRUE, stringsAsFactors = FALSE)

# Create a bar plot showing proportions with black borders
ggplot(race_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
  labs(title = "Proportions of Causes of Death by Category",
       x = "Category",
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
